---
title: "Simulation and Game Analysis"
description: >
  Self-play between policies, opening statistics with Wilson intervals,
  per-ply analysis (win probability, regret, missed wins/blocks, turning
  points), and behavioral profiles for comparing playing styles.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Simulation and Game Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5, fig.height = 4,
  out.width = "100%"
)
```

```{r setup}
library(tesseractR)
```

> **Prerequisite.** This vignette assumes the API from
> [Getting Started with tesseractR](tesseractR.html). If you'd rather just
> *play*, the four-part tutorial starts at
> [The rules of 4D tic-tac-toe](https://r-heller.github.io/tesseractR/articles/play-01-rules.html).

`tesseractR` is also a *study tool*: a single evaluation core
(`tsr_evaluate()` / `tsr_win_prob()` / `tsr_rate_moves()`) powers self-play
simulation, per-ply game analysis, and aggregate behavioral profiles.
Everything below routes through that core — no parallel scoring logic.

## A single self-play game

```{r}
g <- tsr_play_game("greedy", "random", seed = 7L)
g
```

The `tsr_game` object records the move sequence, who moved on each ply, the
final board, the outcome, and the policy labels. Replaying the moves on a
fresh board reproduces `final_board` exactly.

```{r, fig.alt = "Final position of one greedy-vs-random game"}
tsr_plot(g$final_board)
```

## Per-ply analysis

`tsr_analyze_game()` walks the game and for each ply records: win-probability
before/after, the move's `delta`, the best alternative, regret, and flags
for missed wins, missed blocks, and turning points
(`abs(delta) >= 0.15`).

```{r}
analysis <- tsr_analyze_game(g)
tail(analysis, 6)
```

```{r, fig.alt = "Win-probability trajectory across plies, turning points marked"}
tsr_plot_winprob(analysis)
```

```{r}
tsr_turning_points(analysis)
tsr_game_summary(g)
```

## Self-play in bulk

`tsr_simulate()` runs many games and returns a tibble with the
first-move position broken into its four coordinates — the substrate for
opening statistics.

> **Performance.** Pure-R self-play is throughput-bounded; the heavy hot
> paths (`tsr_evaluate`, legal-move generation) are flagged for future
> Rcpp replacement. Meaningful opening statistics need thousands of games;
> the example below is deliberately tiny so the vignette builds in seconds.

```{r}
sim <- tsr_simulate("random", "random", n_games = 12L,
                    seed = 1L, verbose = FALSE)
sim
```

```{r}
opening <- tsr_opening_stats(sim, by = "first_move")
head(opening, 8)
```

The opening-statistics tibble carries a Wilson 95% confidence interval on
`win_rate_x` — small `n_games` means wide intervals. A few hundred games
per opening would be needed to draw real conclusions; this vignette uses a
deliberately tiny sample so it builds in seconds.

## Policy behavioral profile

`tsr_behavior_profile()` aggregates per-ply analyses across many games into
a single behavioral signature for one side. Use `tsr_compare_profiles()` to
stack profiles side by side.

```{r}
games_r <- lapply(1:3, function(s) tsr_play_game("random", "random", seed = s))
games_g <- lapply(1:3, function(s) tsr_play_game("greedy", "random", seed = s))
p_random <- tsr_behavior_profile(games_r, side = 1L, label = "X random")
p_greedy <- tsr_behavior_profile(games_g, side = 1L, label = "X greedy")
tsr_compare_profiles(p_random, p_greedy)
```

Read the rows side-by-side: `accuracy` is the share of plies that matched
the engine's top move; `blunder_rate` is the share with regret above the
documented threshold; `aggression` is the share of moves that opened a new
3-in-line threat for the side.

## Where this all meets the app

The same `tsr_win_prob()` powers the app's live win-probability gauge, the
same `tsr_rate_moves()` powers the per-cell rating overlay, and the same
`tsr_analyze_game()` powers the post-game analysis tab — so a number you
see in the UI matches exactly what these functions produce from the console.
