---
title: "Getting Started with tesseractR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with tesseractR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5, fig.height = 6.5,
  out.width = "100%"
)
```

```{r setup}
library(tesseractR)
```

## What is 4D tic-tac-toe?

`tesseractR` plays tic-tac-toe on a four-dimensional **4×4×4×4** hypercube —
256 cells indexed by `(i, j, k, l)`, each coordinate in `0:3`. A *line* is any
four cells whose successive differences equal a constant non-zero direction
vector in `{−1, 0, 1}^4`. After collapsing reversed directions there are
**520 winning lines** by the standard formula
`((n + 2)^d − n^d) / 2` evaluated at `d = n = 4`.

The geometry is the only thing that's harder than ordinary tic-tac-toe.
Everything else (engine, AI, visualization, analysis) is the usual surface,
just generalised over the 256-cell board.

## A first game

Each cell has a unique linear index in `1:256` derived from
`idx = 1 + i + 4j + 16k + 64l`. You can call `tsr_move()` with explicit
coordinates or with `cell =`:

```{r}
b <- tsr_new_board()
b <- tsr_move(b, 0L, 0L, 0L, 0L)    # X at (0,0,0,0) = cell 1
b <- tsr_move(b, cell = 17L)        # O at cell 17
b <- tsr_move(b, 1L, 0L, 0L, 0L)    # X at (1,0,0,0) = cell 2
print(b)
```

```{r, fig.alt = "Mid-game tesseractR board on the 4x4 of 4x4 layout"}
tsr_plot(b)
```

The plot draws the hypercube as a 4×4 outer grid (indexed by `(k, l)`) where
each cell contains an inner 4×4 board (indexed by `(i, j)`). Player X is the
blue mark; player O is orange (Okabe–Ito).

## Detecting a win

Win detection scans the cached 520-line table against the board state.
Continuing the game until X completes an axis-aligned line:

```{r}
b <- tsr_move(b, cell = 18L)        # O
b <- tsr_move(b, 2L, 0L, 0L, 0L)    # X
b <- tsr_move(b, cell = 19L)        # O
b <- tsr_move(b, 3L, 0L, 0L, 0L)    # X completes 1-2-3-4
tsr_check_win(b)
tsr_winning_line(b)
tsr_status(b)
```

```{r, fig.alt = "Full board with the winning line highlighted"}
tsr_plot(b)
```

The slice view fixes one axis at one value and renders the resulting 4×4×4
cube as a strip — handy for inspecting a plane in which an otherwise
scattered diagonal reads as a straight line:

```{r, fig.height = 2.4, fig.alt = "Slice along l = 0 containing the winning line"}
tsr_plot_slice(b, axis = "l", at = 0L)
```

## Legal moves and undo

`tsr_legal_moves()` is always a type-stable integer vector (length 0 on a
finished or full board). `tsr_undo()` rewinds moves immutably:

```{r}
length(tsr_legal_moves(tsr_new_board()))
length(tsr_legal_moves(b))

# Undo to a non-terminal position
b_pre <- tsr_undo(b, n = 1L)
tsr_status(b_pre)
```

## Playing the AI

`tsr_ai_move()` returns the chosen linear cell index. Difficulty `1` always
takes an immediate win and blocks an immediate threat; higher difficulties
search further ahead (depth-limited negamax with alpha-beta). Search is
exponential in pure R — depth 4 may take several seconds; Rcpp acceleration
is a planned future enhancement.

```{r}
b <- tsr_new_board()
b <- tsr_move(b, cell = 1L)              # human X
ai_o <- tsr_ai_move(b, difficulty = 1L)  # AI O
b <- tsr_move(b, cell = ai_o)
print(b)
```

## Evaluating positions

`tsr_evaluate()` returns a unitless raw score; positive favours `player`.
`tsr_win_prob()` calibrates the score into `[0, 1]` via a logistic mapping
whose coefficients are fitted from self-play data (S6) and shipped as
internal package data.

```{r}
b <- tsr_new_board()
b <- tsr_move(b, cell = 1L); b <- tsr_move(b, cell = 17L)
b <- tsr_move(b, cell = 2L)
tsr_evaluate(b, player = 1L)
tsr_win_prob(b, player = 1L)
```

`tsr_rate_moves()` is the per-move best-next-move surface used by the AI,
the app's overlay, and the analysis routines:

```{r}
head(tsr_rate_moves(b), 5)
```

## The interactive app

`tsr_run_app()` launches the Shiny version with a live move-rating overlay,
a win-probability gauge, and a game-analysis panel. The Full / Slice (2D) /
Cube (3D) view toggle lets you change perspective without changing state.

```r
tsr_run_app(difficulty = 2L)
```

See **"Simulation and Game Analysis"** for self-play and post-game analytics.
