---
title: "Getting Started with tesseractR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tesseractR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 6
)
```

```{r setup}
library(tesseractR)
```

## What is 4D tic-tac-toe?

`tesseractR` plays tic-tac-toe on a four-dimensional 4×4×4×4 hypercube — 256
cells indexed by `(i, j, k, l)` with each coordinate in `0:3`. A line is any
four cells whose successive differences equal a constant non-zero direction
vector in `{−1, 0, 1}^4`. After deduplicating reversed directions there are
**520 winning lines** by the standard `((n+2)^d − n^d) / 2` formula for
`d = n = 4`.

## A first game

```{r}
b <- tsr_new_board()
b <- tsr_move(b, 0L, 0L, 0L, 0L)   # X at (0,0,0,0) = cell 1
b <- tsr_move(b, cell = 17L)       # O
b <- tsr_move(b, 1L, 0L, 0L, 0L)   # X at (1,0,0,0) = cell 2
print(b)
```

```{r, fig.alt = "Mid-game tesseractR board"}
tsr_plot(b)
```

## Detecting a win

We continue the game until X completes an axis-aligned line:

```{r}
b <- tsr_move(b, cell = 18L)       # O
b <- tsr_move(b, 2L, 0L, 0L, 0L)   # X
b <- tsr_move(b, cell = 19L)       # O
b <- tsr_move(b, 3L, 0L, 0L, 0L)   # X completes 1-2-3-4
tsr_check_win(b)
tsr_winning_line(b)
```

```{r, fig.alt = "Slice view highlighting the in-plane winning line"}
tsr_plot_slice(b, axis = "l", at = 0L)
```

## Playing the AI

`tsr_ai_move()` returns the chosen linear cell index. Difficulty `1` always
takes an immediate win and blocks an immediate threat. Higher difficulties
look further ahead (search is exponential in pure R — depth 4 may take
several seconds).

```{r}
b <- tsr_new_board()
b <- tsr_move(b, cell = 1L)        # human X
ai_cell <- tsr_ai_move(b, difficulty = 1L)
b <- tsr_move(b, cell = ai_cell)   # AI O
ai_cell
```

## Evaluating positions

```{r}
b <- tsr_new_board()
b <- tsr_move(b, cell = 1L); b <- tsr_move(b, cell = 17L)
b <- tsr_move(b, cell = 2L)
tsr_win_prob(b, player = 1L)
head(tsr_rate_moves(b))
```

## The app

`tsr_run_app()` launches the interactive Shiny version with a live
move-rating overlay and a game-analysis panel. See the
"Simulation and Game Analysis" vignette for self-play and analytics.
