---
title: "4. Playing from the R console"
description: >
  Play tesseractR directly in R: create a board, make and undo moves, list legal
  moves, ask the AI for a move, evaluate positions, and plot the result.
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 7,
  out.width = "100%",
  dpi = 120
)
```

```{r setup}
library(tesseractR)
```

> **Tutorial page 4 of 4.**
> Previous: [« Seeing in four dimensions](play-03-strategy.html) ·
> Back to the start: [The rules](play-01-rules.html)

The app is the friendly front end, but the whole game is scriptable. This page is
the console version of the same moves.

## A board and a move

A game starts from an empty board. X is always to move first.

```{r board}
b <- tsr_new_board()
b
```

Make a move with either explicit coordinates or a linear `cell` index — the two
calls below are equivalent ways of naming cell 1, `(0, 0, 0, 0)`:

```{r move}
b <- tsr_move(b, 0L, 0L, 0L, 0L)   # X at (0,0,0,0)
b <- tsr_move(b, cell = 17L)        # O at cell 17 = (0,0,1,0)
b <- tsr_move(b, 1L, 0L, 0L, 0L)   # X at (1,0,0,0)
b
```

`tsr_status()` is the machine-readable summary the app shows in its sidebar:

```{r status}
tsr_status(b)
```

Made a mistake? `tsr_undo()` rolls back one or more plies:

```{r undo}
b <- tsr_undo(b, n = 1L)    # take back O's reply... err, the last move
tsr_status(b)$n_moves
```

## Listing the legal moves

`tsr_legal_moves()` returns the cell indices that are still empty:

```{r legal}
moves <- tsr_legal_moves(b)
length(moves)
head(moves, 8)
```

## Asking the AI for a move

`tsr_ai_move()` returns the cell the negamax engine would play, with a tunable
search depth via `difficulty` (1–4):

```{r ai}
ai_cell <- tsr_ai_move(b, difficulty = 2L)
ai_cell
b <- tsr_move(b, cell = ai_cell)
tsr_status(b)
```

## A complete game loop

Putting it together: a short game where *you* play a fixed opening and the AI
answers each move. (In a real session you would choose moves interactively; here
the human side is scripted so the article is reproducible.)

```{r loop}
set.seed(1)
b <- tsr_new_board()
my_moves <- c(1L, 2L, 3L, 86L, 171L)   # a scripted "human" (player X)

for (cell in my_moves) {
  if (tsr_status(b)$is_over) break
  legal <- tsr_legal_moves(b)
  if (!(cell %in% legal)) cell <- legal[1]            # fall back if AI took it
  b <- tsr_move(b, cell = cell)                       # X (you)
  if (tsr_status(b)$is_over) break
  b <- tsr_move(b, cell = tsr_ai_move(b, difficulty = 2L))  # O (AI)
}

tsr_status(b)
```

```{r loop-plot, fig.alt = "The board after a short scripted game between a human opening and the AI."}
tsr_plot(b)
```

## Evaluating a position

Three functions read a position, all sharing one evaluation core:

```{r eval}
tsr_evaluate(b)                       # raw score for the side to move
tsr_win_prob(b, method = "heuristic") # squashed to a 0..1 win probability
```

`tsr_rate_moves()` scores *every* legal move at once and is the engine behind the
app's overlay. The columns `is_best`, `is_winning`, and `is_blocking` are the
quickest way to find the move that matters:

```{r rate}
rated <- tsr_rate_moves(b, method = "heuristic")
head(rated[order(-rated$win_prob),
           c("cell", "i", "j", "k", "l", "win_prob", "is_best")], 5)
```

## Where to go next

* Run the same game interactively with `tsr_run_app()` — see
  [Playing in the app](play-02-app.html).
* Simulate many games and profile playing styles in
  [Simulation and Game Analysis](analysis.html).
* Browse every function in the [reference index](../reference/index.html).

> Back to the start: [The rules of 4D tic-tac-toe »](play-01-rules.html)
