---
title: "1. The rules of 4D tic-tac-toe"
description: >
  How the 4x4x4x4 hypercube board is laid out, what counts as a winning line,
  and how to read the board plot. The first page of the playing tutorial.
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 7,
  out.width = "100%",
  dpi = 120
)
```

```{r setup}
library(tesseractR)
```

> **This is a four-part tutorial on how to *play* tesseractR.**
> You are on page 1. Next: [Playing in the app »](play-02-app.html)

## From a square to a tesseract

Ordinary tic-tac-toe is played on a 3×3 square: a line of three wins. `tesseractR`
lifts the same idea into **four dimensions**. The board is a **4×4×4×4
hypercube** — a *tesseract* — with **256 cells**. A line of **four** marks wins.

Every cell has four coordinates `(i, j, k, l)`, each in `0:3`:

* `i`, `j` — position *inside* one little 4×4 board,
* `k`, `l` — which little board, within the outer 4×4 grid of boards.

Each cell also has a single linear index in `1:256`, which is how you refer to a
cell with `cell =`:

$$\text{idx} = 1 + i + 4j + 16k + 64l.$$

So `(0, 0, 0, 0)` is cell `1`, and `(0, 0, 0, 1)` is cell `r 1 + 16*4`.

## Reading the board

`tsr_plot()` draws the hypercube as a **4×4 outer grid of 4×4 boards**. The outer
grid is indexed by `(k, l)`; each inner board by `(i, j)`. An empty board:

```{r empty, fig.alt = "An empty tesseractR board: a 4x4 grid of empty 4x4 boards."}
tsr_plot(tsr_new_board())
```

Player **X** is the blue mark, player **O** is the orange mark (the colour-blind
friendly Okabe–Ito palette). X always moves first.

## What is a winning line?

A **line** is any four cells whose successive coordinate differences equal a
constant non-zero step in `{-1, 0, 1}^4`. That single rule covers a lot of
geometry:

* **straight rows** inside one little board (vary `i`, hold `j, k, l`),
* lines that march **across boards** (vary `k` or `l`),
* and full four-dimensional **diagonals** that change all four coordinates at
  once.

After collapsing each line and its reverse, there are

$$\frac{(n+2)^d - n^d}{2} = \frac{6^4 - 4^4}{2} = 520$$

winning lines for `d = n = 4` — far more than the 8 lines of the 3×3 game, which
is exactly what makes four dimensions hard to hold in your head.

Here is a finished game where X has completed a line of four. When the game is
over, `tsr_plot()` rings the winning cells in pink:

```{r win, fig.alt = "A finished board where player X has four marks in a row, ringed in pink."}
b <- tsr_new_board()
moves <- list(
  c(0, 0, 0, 0), c(0, 1, 0, 0),   # X, O
  c(1, 0, 0, 0), c(1, 1, 0, 0),   # X, O
  c(2, 0, 0, 0), c(2, 1, 0, 0),   # X, O
  c(3, 0, 0, 0)                    # X completes i = 0..3 at j=k=l=0
)
for (m in moves) b <- tsr_move(b, m[1], m[2], m[3], m[4])

tsr_status(b)
tsr_plot(b)
```

The four X's sit in the **top-left** little board (that is the `k = 0, l = 0`
board), along its top row (`j = 0`), as `i` runs `0 → 3`. The pink rings mark
exactly those four cells.

## Where lines hide

Most winning lines are *not* a tidy row inside one board — they jump between
boards. To make those read as straight lines, `tsr_plot_slice()` fixes one axis
and shows the resulting 3D cube as a strip of four boards. We return to this in
[Seeing in four dimensions](play-03-strategy.html); for now, just note that the
same board can be sliced four different ways:

```{r slice, fig.width = 9, fig.height = 3, fig.alt = "A slice view of the board: four 4x4 boards in a row."}
tsr_plot_slice(b, axis = "l", at = 0)
```

## Recap

* 256 cells, coordinates `(i, j, k, l)` in `0:3`; cell index `1 + i + 4j + 16k + 64l`.
* X moves first; four in a line wins; there are **520** winning lines.
* `tsr_plot()` shows the full board, `tsr_plot_slice()` shows one 3D slice.

> Next: [Playing in the app »](play-02-app.html)
